---
title: "Flight Delays - Final project"
output: html_notebook
---

# Flight Delays - Final project. 

## 5. Analysis
### Step 1: Flight delays - quantifying delays at Newark and other NY Airports

```{r}
library(tidyverse)
library(here)
library(GGally)
library(modelr)
library(leaflet)
library(caret)
library(ranger)
library(janitor)
library(pROC)
library(broom)

```


```{r}
flights_clean <- read_csv(here("clean_data/flight_clean.csv"))
```



```{r}
flights_clean %>% 
  filter(origin == "EWR") %>% 
  group_by(delay_dep) %>% 
  count() %>% 
  ggplot()+
  geom_col(aes(x = reorder(delay_dep, n), y = n), fill = "steelblue")+
  coord_flip()+
  theme_classic()+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  labs(
    x = "Flight delay",
    y = "Number of flights",
    title = "Number of flights delayed from Newark Airport in 2017"
  )+
  theme(plot.title = element_text(hjust = 1))
```

```{r}
# calculate the proportions and numbers of delayed flights and by how much
ewr_flights <- flights_clean %>% 
  filter(origin == "EWR") 

ewr_flights %>% 
  group_by(delay_dep) %>% 
  summarise(number = n(),
            prop = n()/nrow(ewr_flights)*100)
```


```{r}
# compare flight delayed across each airport in New York
flights_clean %>% 
  group_by(delay_dep, origin) %>% 
  summarise(count = n()) %>% 
  ggplot()+
  geom_col(aes(x = origin, y = count, fill = delay_dep), position = "fill")+
  theme_classic()+
  labs(
    x = "Airport",
    y = "Proportion of flights",
    title = "Comparison of Flight Delays from New York Airports in 2017",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))
```

## Data Wrangling

Initial explorations of data showed a lot of missing weather data points. Decided to deal with these by imputing the mean from that day. Also, as the airports are in a similar location, if there were no data points for a whole day, then the means were taken from the other NY airports. 


```{r}
all_joined <- read_csv("clean_data/all_joined.csv")
```


```{r}
# impute missing weather data to include daily means

all_joined_imputed <- all_joined %>% 
  group_by(date, origin) %>% 
  mutate(across(.cols = temp:wind_speed, .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)),
         across(.cols = wind_gust:visib, .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)))

```

```{r}
# impute data to include monthly means
all_joined_imputed <- all_joined_imputed %>% 
  group_by(month, origin) %>% 
  mutate(temp = if_else(is.na(temp), mean(temp, na.rm = TRUE), temp))
```

```{r}
# impute data to include monthly means
all_joined_imputed <- all_joined_imputed %>% 
  group_by(month, origin) %>% 
  mutate(across(.cols = c(temp, precip, pressure, humid, dewp), .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)))
```


```{r}
# include wind_speed category
all_joined_imputed <- all_joined_imputed %>% 
  mutate(wind_class = case_when(wind_speed <= 24 ~"Breeze",
                                wind_speed <= 31 ~"Strong breeze",
                                wind_speed <= 38 ~ "Near gale",
                                wind_speed > 38 ~ "Gale"),
         .after = wind_speed)

```


```{r}
# introduce a wind_direction category variable from wind_dir
all_joined_imputed <- all_joined_imputed %>% 
  mutate(wind_direction = case_when(wind_dir <= 23 ~ "North",
                                    wind_dir <= 68 ~ "Northeast",
                                    wind_dir <= 113 ~ "East",
                                    wind_dir <= 158 ~ "Southeast",
                                    wind_dir <= 203 ~ "south",
                                    wind_dir <= 248 ~ "southwest",
                                    wind_dir <= 293 ~ "West",
                                    wind_dir <= 336 ~ "northwest",
                                    wind_dir > 336 ~ "North"),
         .after = wind_dir,
         delayed = as.logical(delayed),
         delayed_1hr_more = as.logical(delayed_1hr_more),
         airport = if_else(origin == "EWR", "EWR", "other NY airport"))

head(all_joined_imputed)

```

### Step 2: Impact of weather on flight delays


```{r message=FALSE}
# carry out correlation plots between weather types and delayed flights
all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  ungroup() %>% 
  select(delayed, temp, pressure, visib, dewp, wind_speed, wind_gust, wind_dir, humid) %>% 
  ggpairs()
```


```{r message=FALSE}
# carry out correlation plots between weather types and length of flight delay
all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  ungroup() %>% 
  select(dep_delay, temp, pressure, visib, dewp, wind_speed, wind_gust, wind_dir, humid) %>% 
  ggpairs()
```
```{r message=FALSE}
# carry out correlation plots between weather types and delayed flights
all_joined_imputed %>% 
  ungroup() %>% 
  select(delayed, delay_dep, wind_class, wind_direction) %>% 
  ggpairs()
```

#### Temperature
```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = temp, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Temperature (F)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```

#### Wind Gusts

```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = wind_gust, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Wind gusts (mph)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```
#### Precipitation

```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = precip, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Precipitation (inches)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```

#### Wind Speed

```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = wind_speed, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Wind speed (mph)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```

#### Comparison of average flight delays for each weather time, and by airport


```{r}
# comparison of impact of weather type on flight delays across all NY airports
all_joined_imputed %>% 
  group_by(wind_direction, delayed, airport) %>% 
  summarise(mean = mean(wind_speed)) %>% 
  ggplot(aes(x = wind_direction, y = mean, fill = delayed))+
  geom_col(position = "dodge")+
  theme_classic()+
  facet_wrap(~airport)+
  coord_flip()+
  labs(
    x = "Wind speed (mph)",
    y = "Delayed flights",
    fill = "Departure\ndelay",
    title = "Effect of Wind Direction on Flight Delays from New York Airports"
  )+
  scale_fill_manual(values = c("#fdae61", "#2c7bb6"))+
  theme(axis.text=element_text(size=16))+
  theme(text = element_text(size = 14))+
  theme(plot.title = element_text(size = 16, hjust = 0.4))
```


```{r}
# comparison of all weather types effect on flight delays at Newark airport
avg_table <- all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  group_by(delayed) %>% 
  summarise(Temperature = mean(temp),
            #pressure = mean(pressure, na.rm = TRUE),
            Visibility = mean(visib, na.rm = TRUE),
            Humidity = mean(humid, na.rm = TRUE),
            #precip = mean(precip, na.rm = TRUE),
            "Wind speed" = mean(wind_speed, na.rm = TRUE),
            "Wind gust" = mean(wind_gust, na.rm = TRUE))

avg_table_pivot <- avg_table %>% 
  pivot_longer(cols = c(Temperature:"Wind gust"), names_to = "weather", values_to = "mean")

avg_table_pivot %>% 
  ggplot(aes(x = weather, y = mean, fill = delayed))+
  geom_col(position = "dodge")+
  theme_classic()+
  labs(
    y = "Mean weather",
    title = "Impact of weather on flight delays from Newark airport",
    fill = "Delayed"
  )+
  theme(axis.text=element_text(size=12))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  scale_fill_manual(values = c("#2c7bb6","#fdae61", "#abd9e9",  "#d7191c" ))
```



```{r}
# compare destribution of average wind speed on flight delay in different airports
all_joined_imputed %>% 
  filter(delayed == TRUE) %>% 
  group_by(wind_speed, airport) %>% 
  summarise(mean = mean(dep_delay)) %>% 
  ggplot(aes(y = mean, x = wind_speed, colour = airport))+
  geom_point()+
  facet_grid(~airport)+
   theme_classic()+
  labs(
    x = "Wind speed (mph)",
    y = "Mean delay (minutes)",
    title = "Average Flight Delay with Wind Speed at New York Airports"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  theme(legend.position="none")
  
```


### Step 3: Airplot location

```{r}
# create new table with airport information, flight delays, destance
airport_table <- all_joined %>% 
  filter(origin == "EWR") %>% 
  group_by(name.y) %>% 
  summarise(count = n(),
            delayed_flights = sum(delayed),
            prop_delayed = delayed_flights/count,
            airtime = mean(air_time),
            distance = mean(distance)) %>% 
  arrange(desc(prop_delayed))
```



```{r}
# add in additional information for latitude and long missing. Easier to do this in excel.

airports <- read_csv(here("clean_data/airports.csv"))

airport_table_latlong <- airport_table %>% 
  left_join(airports, by = c("name.y" ="name"))

write_csv(airport_table_latlong, "clean_data/airport_table_latlong.csv", append = FALSE)
```


```{r}
airport_latlong_updated <- read_csv("clean_data/airport_table_latlong_updated.csv")
```



```{r}
# plot correlation between how busy an airport is and the proportion of delayed flights
airport_latlong_updated %>% 
  ggplot(aes(x = count, y = prop_delayed))+
  geom_point(colour = "steelblue")+
  labs(
    x = "Number of flights Airport",
    y = "Proportion of delayed flights",
    title = "Proportion of Delayed flights compared to Total Flights by Airport"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.7))
```



```{r}
# add in category from quartiles for how good/bad the proportion of flight delays are. Also, create a colour palette for groups
airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(delay_class = case_when(prop_delayed <= 0.31 ~ "1",
                                 prop_delayed <= 0.3738 ~ "2",
                                 prop_delayed <= 0.4143 ~ "3",
                                 prop_delayed > 0.4143 ~ "4")) 

pal <- colorFactor(
  palette = c('green', 'blue', 'orange', 'red'),
  domain = airport_latlong_updated$delay_class
)
size <- c()
```


```{r}
# categorise how busy an airport is. Include a reasonable size range for leaflet
airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(count_group = case_when(count < 500 ~"<500",
                                 count < 1000 ~ "<1000",
                                 count < 2000 ~ "<2000",
                                 count < 3000 ~ "<3000",
                                 count < 4000 ~ "<4000",
                                 count < 5000 ~ "<5000",
                                 count < 6000 ~ "<6000",
                                 count < 7000 ~ "<7000",
                                 count > 7000 ~ ">7000"))

airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(count_sqrt = sqrt(count)/3)
```

```{r}
# geospatial mapping of airport locations, flight delays and how busy the airports are
airport_latlong_trim <- airport_latlong_updated %>% 
  mutate(across(c(count, delayed_flights, lat, lon), as.numeric)) %>% 
  filter(!is.na(lat)) %>% 
  filter(!is.na(lon))

leaflet(airport_latlong_trim) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~lon,
                   lat = ~lat,
                   weight = 1,
                   radius = ~airport_latlong_updated$count_sqrt,
                   color = ~pal(delay_class)) %>% 
  addLegend('bottomright', pal = pal, values = airport_latlong_updated$delay_class,
            title = 'Proportion of<br>flights delayed',
            opacity = 1)
```


```{r}
# correlate the number of delayed flights to how busy an airport is
airport_latlong_updated %>% 
  ggplot(aes(x = count, y = delayed_flights))+
  geom_point()+
  geom_smooth()+
  theme_classic()+
  labs(
    x = "Number of flights to Airport",
    y = "Number of delayed flights",
    title = "Number of Delayed flights compared to Total Flights by Airport"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.4))
```



```{r}
# correlate the proportion of delayed flights to the distance to the destination
all_joined_imputed %>% 
  filter(delayed == TRUE,
         origin == "EWR") %>% 
  group_by(distance, delayed) %>% 
  summarise(mean_delay = mean(dep_delay)) %>% 
  ggplot(aes(x = distance, y = mean_delay))+
  geom_point(colour = "steelblue")+
  theme_classic()+
  labs(
    x = "Distance of flight",
    y = "Mean delay (minutes)",
    title = "Average time delay versus distance of delayed flight from Newark Airport"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  ylim(0, 100)

```

### Step 4: Aircraft and airline analysis

```{r}
flights_weather_planes_airlines <- read_csv("clean_data/flights_weather_planes_airlines.csv")
```

```{r}
# convert variables to integers for plots
flights_weather_planes_airlines <- flights_weather_planes_airlines %>% 
  mutate(delayed = as.integer(delayed),
         delayed_1hr_more = as.integer(delayed_1hr_more))
```

```{r}
# Remove extra string in text variables
flights_weather_planes_airlines_regex <- flights_weather_planes_airlines %>% 
  mutate(name = str_remove(name, " Inc."),
         name = str_remove(name, " Co"))
```


```{r}
# plot percentage of flight delays for each airline
flights_weather_planes_airlines_regex %>% 
  group_by(name) %>% 
  summarise(count = n(),
            mean_delay = mean(dep_delay),
            perc_flights_delayed = sum(delayed)/count*100,
            perc_flights_delayed_hr = sum(delayed_1hr_more)/count*100) %>% 
  arrange(desc(perc_flights_delayed)) %>% 
  ggplot(aes(x = reorder(name, perc_flights_delayed), y = perc_flights_delayed, ))+
  geom_col(fill = "steelblue")+
  coord_flip()+
  theme_classic()+
  labs(
    y = "Percentage flights delayed",
    title = "Percentage Flight Delays by Airline"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(axis.title.y=element_blank())+
  theme(plot.title = element_text(size = 20, hjust = -0.4))
```

### Step 5: Time of year/day

#### Correlation between departure delay and arrival delay

```{r}
# correlation between departure time and arrival time for delayed flights. ie do delayed flights catch up time during journey. 
flights_clean %>% 
  filter(delayed == "TRUE") %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  ggplot(aes(x = dep_delay, y = arr_delay))+
  geom_point()+
  geom_smooth()
```


```{r}
# plot flight delays by month
all_joined_imputed%>% 
  filter(origin == "EWR") %>% 
  group_by(month, delay_dep) %>% 
  summarise(mean_delay = mean(dep_delay),
            count = n())%>% 
  ggplot()+
  geom_col(aes(x = month, y = count, fill = factor(delay_dep, levels = c("> 1hr delay", "< 1hr delay","< 15 minutes delay",  "Early or on-time"))), position = "fill")+
  theme_classic()+
  labs(
    x = "Month",
    y = "Proportion of delayed flights",
    title = "Proportion of Delayed Flights by Month from Newark",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  scale_x_continuous(breaks=seq(1,12,1))
```


```{r}
# plot flight delays by time of the day
all_joined_imputed%>% 
  filter(origin == "EWR") %>% 
  group_by(hour, delay_dep) %>% 
  summarise(mean_delay = mean(dep_delay),
            count = n())%>% 
  ggplot()+
  geom_col(aes(x = hour, y = count, fill = factor(delay_dep, levels = c("> 1hr delay", "< 1hr delay","< 15 minutes delay",  "Early or on-time"))), position = "fill")+
  theme_classic()+
  labs(
    x = "Hour",
    y = "Proportion of delayed flights",
    title = "Flight Delay by Hour from Newark Airport",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 1))
  
```



```{r}
# compare flight delays by hour across all three New York airports
flights_clean %>% 
  group_by(hour,origin) %>% 
  summarise(mean_delay = mean(dep_delay)) %>% 
  ggplot(aes(x = hour, y = mean_delay, colour = origin))+
  geom_point()+
  geom_line()+
  theme_classic()+
  labs(
    x = "Hour",
    y = "Average flight delay (minutes)",
    title = "Average Flight Delay by Hour from New York Airports",
    fill = "Airport"
  )+
  scale_colour_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 1))+
  scale_x_continuous(breaks = seq(0, 24, 2))
```

## 6. Model building
### Predictions for flight delays using Decision tree

```{r}
ewr_fit <- all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  ungroup() 
  

ewr_fit <- ewr_fit %>% 
  mutate(time_of_day = case_when(hour <= 6 ~ "night",
                                 hour <= 12 ~ "morning",
                                 hour <= 18 ~ "afternoon",
                                 hour <= 24 ~ "evening"),
         visibility = if_else(visib >= 5, "good", "poor"))

ewr_fit_dt <- ewr_fit %>% 
  select(delayed, month, time_of_day, distance, wind_direction, wind_class, temp, visibility, wind_gust, precip) %>% 
  mutate(delayed = as.factor(delayed)) %>% 
  na.omit()
ewr_fit_dt
```





```{r}
# test train split set up
n_data <- nrow(ewr_fit_dt)

test_index <- sample(1:n_data, size = n_data*0.2)

ewr_test <- slice(ewr_fit_dt, test_index)

ewr_train <- slice(ewr_fit_dt, -test_index)
```

```{r}
ewr_test %>% 
  janitor::tabyl(delayed)
```
```{r}
ewr_train %>% 
  janitor::tabyl(delayed)
```

```{r}
# fit data to tree
ewr_fit_tree <- rpart(
  formula = delayed ~.,
  data = ewr_train,
  method = "class"
)

rpart.plot(ewr_fit_tree,
           yesno = 2,
           fallen.leaves = TRUE,
           faclen = 6,
           digits = 2, 
           type = 4,
           extra = 101)
```
```{r}
rpart.rules(ewr_fit_tree, cover = TRUE)
```



```{r}
# add predictions to test the model
ewr_test_pred <- ewr_test %>% 
  add_predictions(ewr_fit_tree, type = "class")
```


```{r}
ewr_test_pred %>% 
  select(time_of_day, month, delayed, pred)  
```

```{r}
# confusion matrix to evaluate model
conf_mat <- ewr_test_pred %>% 
  conf_mat(truth = delayed, estimate = pred)

conf_mat
```

```{r}
accuracy <- ewr_test_pred %>% 
  accuracy(truth = delayed, estimate = pred)

accuracy
```


```{r}
confusionMatrix(ewr_test_pred$pred, ewr_test_pred$delayed)
```
The conclusion from this decision tree is that it only takes in one variable (hour) as this is the only variable that means the defined threshold for convergence. This reinforces previous analysis that there is no difference in the delays in flights with different weather. 

### Predictions for flight delays using Decision tree


```{r}
ewr_fit %>% 
  summary
```

```{r}
ewr_trim <- ewr_fit %>% 
  select(month, day, delayed, air_time, distance, hour, temp, wind_direction, wind_speed, wind_class, wind_gust, visibility, year.y, type, manufacturer, model, engines, seats, engine, name.x, lat, lon, alt, tz) %>% 
  na.omit()


ewr_trim 
```


```{r}
rf_classifier <- ranger(delayed ~ .,
                       data = ewr_trim, 
                       importance = "impurity",
                       num.trees = 1000,
                       mtry = 2, 
                       min.node.size = 5)

rf_classifier
```

```{r}
# view varibles with importance value
sort(importance(rf_classifier))
```

From this we see that hour, airtime, day, minute and year.y(ie year of plane) are the most important variables, followed by wind_gust, wind_speed, month, wind_direction, temperature, model, lat, distance, altitude, longitude, seats, name.x, manufacturer, visibility, tz, engine, wind_class, type and engines. 



### Predictions for flight delays using logistical regression

Based on the delayed variable (True/false) and single variable for hour. 

```{r}
# set up the dataset

all_joined_model <- all_joined_imputed %>% 
  select(date, time, delayed, hour)
```


```{r}
n_data <- nrow(all_joined_model)

delay_index <- sample(1:n_data, size = n_data*0.2)

delay_test <- slice(all_joined_model, test_index)

delay_train <- slice(all_joined_model, -test_index)
```

```{r}
delay_test %>% 
  janitor::tabyl(delayed)
```
```{r}
delay_train%>% 
  janitor::tabyl(delayed)
```


```{r}
delayed_1log_reg <- glm(delayed ~ hour, 
                             data = delay_train,
                             family = binomial(link = "logit"))

delayed_1log_reg
```


```{r}
test_1log_reg <- glm(delayed ~ hour,
                     data = delay_test, 
                     family = binomial(link = "logit"))
```

Interpretation

ln(odds_delayed(x)) = b0 + b1 * x

in this case:
ln(odds_delayed(x)) = -2.1996 + (0.1242 * x)

$$              
\ln({odds_{delayed}(x))}=-2.1996 + (0.1242 * (x))

$$

```{r}
# add predictions
predict_log <- tibble(hour = seq(0, 24, 0.5)) %>% 
  add_predictions(delayed_1log_reg, 
                  type = "response")

predict_log

delay_train %>% 
  ggplot(aes(x = hour, y = as.integer(delayed))) +
  geom_jitter(position = position_jitter(h = 0.1), shape = 1)+
  geom_line(data = predict_log,
            aes(y = pred),
            col = "red")+
  labs(title = "Probability of Flight Delay Based on Hour",
       y = "Probability",
       x = "Hour")+
  theme_classic()+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  scale_x_continuous(breaks=seq(0,24,2))
```

```{r}
tidy_out <- clean_names(tidy(delayed_1log_reg))
tidy_out
```

```{r}
tidy_test <- clean_names(tidy(test_1log_reg))
tidy_test
```


```{r}
summary(delayed_1log_reg)
```


```{r}
delay_train_with_1pred <- delay_train %>% 
  add_predictions(delayed_1log_reg, type = "response")

delay_train_with_1pred %>% 
  ggplot(aes(x = hour, y = pred, colour = delayed))+
  geom_line()
```

```{r}
test_with_1pred <- delay_test %>% 
  add_predictions(delayed_1log_reg, type = "response")

test_with_1pred %>% 
   ggplot(aes(x = hour, y = pred, colour = delayed))+
  geom_line()
```


Neither threshold is particularly good!

```{r}
roc_obj_1pred <- delay_train_with_1pred %>% 
  roc(response = delayed,
      predictor = pred)

auc(roc_obj_1pred)
```
```{r}
roc_obj_test1pred <- test_with_1pred %>% 
  roc(response = delayed,
      predictor = pred)

auc(roc_obj_test1pred)
```




```{r}
roc_curve <- ggroc(data = list(
  mod1 = roc_obj_1pred,
  mod2 = roc_obj_test1pred),
                   legacy.axes = TRUE)+
  coord_fixed()+
  labs(x = "FPR = 1 - TNR = 1 - specificity",
          y = "TPR = sensitivity")

roc_curve
```
The simple logistical model with a single predictor (hour of day) shows a better than guess AUC of 0.66. It could be improved by the addition of more predictors, however the p-values show that the variable is significant and can be used to determine the probability (and the odds) of a flight being delayed, based on a single predictor of hour of flight. 




