---
title: "Flight Delays - Final project"
output: html_notebook
---

# Flight Delays - Final project. 

## Introduction and project brief. 

Newark Airport belive that poor weather conditions are causing too many delays and want to invest in improving facilities, so that aircraft can take off in more types of weather. However, they do not fully understand how serious weather related delays are, and are not sure what type of weather they should be most concerned about. As part of investigating the effect of weather you should investigate other factors to understand how important weather is in comparison to them. They also want to understand how they compare to other New York airports.

## Defining the Project. 

From the project brief I identified 4 key questions that will be most important for providing actionable information that Newark Airport can develop. These questions are: 

1. How does weather impact flight delays?
2. How does Newark airport compare to other New York Airports
3. Other factors for delays?
4. Is it possible to build a model which can predict which flights will be delayed?



## Planning

I planned the project with these stages. 


1. Data cleaning (detailed in a separate cleaning script)
2. Data wrangling
3. Initial explorations
4. Analysis
5. Model building
6. Conclusions and insights for action
7. Presentation
8. Further work. 

I roughly allocated a day for each of these tasks, with exception for Analysis section which will be extended to 2-3 days. 



```{r}
library(tidyverse)
library(here)
library(GGally)
library(modelr)
library(leaflet)
library(caret)
library(ranger)
library(janitor)
library(pROC)
library(broom)

```
## Initial Data Exploration - quantifying delays at Newark and other NY Airports

```{r}
flights_clean <- read_csv(here("clean_data/flight_clean.csv"))
```



```{r}
flights_clean %>% 
  filter(origin == "EWR") %>% 
  group_by(delay_dep) %>% 
  count() %>% 
  ggplot()+
  geom_col(aes(x = reorder(delay_dep, n), y = n), fill = "steelblue")+
  coord_flip()+
  theme_classic()+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  labs(
    x = "Flight delay",
    y = "Number of flights",
    title = "Number of flights delayed from Newark Airport in 2017"
  )+
  theme(plot.title = element_text(hjust = 1))
```

```{r}
ewr_flights <- flights_clean %>% 
  filter(origin == "EWR") 

ewr_flights %>% 
  group_by(delay_dep) %>% 
  summarise(number = n(),
            prop = n()/nrow(ewr_flights)*100)
```


```{r}
flights_clean %>% 
  group_by(delay_dep, origin) %>% 
  summarise(count = n()) %>% 
  ggplot()+
  geom_col(aes(x = origin, y = count, fill = delay_dep), position = "fill")+
  theme_classic()+
  labs(
    x = "Airport",
    y = "Proportion of flights",
    title = "Comparison of Flight Delays from New York Airports in 2017",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))
```


## Flight delay by time of year/day

Delay proportion by month to check for seasonal variation


```{r}
all_joined_imputed%>% 
  filter(origin == "EWR") %>% 
  group_by(month, delay_dep) %>% 
  summarise(mean_delay = mean(dep_delay),
            count = n())%>% 
  ggplot()+
  geom_col(aes(x = month, y = count, fill = factor(delay_dep, levels = c("> 1hr delay", "< 1hr delay","< 15 minutes delay",  "Early or on-time"))), position = "fill")+
  theme_classic()+
  labs(
    x = "Month",
    y = "Proportion of delayed flights",
    title = "Proportion of Delayed Flights by Month from Newark",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  scale_x_continuous(breaks=seq(1,12,1))
```
## Data Wrangling

Initial explorations of data showed a lot of missing weather data points. Decided to deal with these by imputing the mean from that day. Also, as the airports are in a similar location, if there were no data points for a whole day, then the means were taken from the other NY airports. 


```{r}
all_joined <- read_csv("clean_data/all_joined.csv")
```


```{r}
all_joined_imputed <- all_joined %>% 
  group_by(date, origin) %>% 
  mutate(across(.cols = temp:wind_speed, .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)),
         across(.cols = wind_gust:visib, .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)))

all_joined_imputed %>% 
  summary
```

```{r}
all_joined_imputed <- all_joined_imputed %>% 
  group_by(month, origin) %>% 
  mutate(temp = if_else(is.na(temp), mean(temp, na.rm = TRUE), temp))
```

```{r}
all_joined_imputed <- all_joined_imputed %>% 
  group_by(month, origin) %>% 
  mutate(across(.cols = c(temp, precip, pressure, humid, dewp), .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)))
```


```{r}
all_joined_imputed <- all_joined_imputed %>% 
  mutate(wind_class = case_when(wind_speed <= 24 ~"Breeze",
                                wind_speed <= 31 ~"Strong breeze",
                                wind_speed <= 38 ~ "Near gale",
                                wind_speed > 38 ~ "Gale"),
         .after = wind_speed)

```


```{r}
all_joined_imputed <- all_joined_imputed %>% 
  mutate(wind_direction = case_when(wind_dir <= 23 ~ "North",
                                    wind_dir <= 68 ~ "Northeast",
                                    wind_dir <= 113 ~ "East",
                                    wind_dir <= 158 ~ "Southeast",
                                    wind_dir <= 203 ~ "south",
                                    wind_dir <= 248 ~ "southwest",
                                    wind_dir <= 293 ~ "West",
                                    wind_dir <= 336 ~ "northwest",
                                    wind_dir > 336 ~ "North"),
         .after = wind_dir,
         delayed = as.logical(delayed),
         delayed_1hr_more = as.logical(delayed_1hr_more),
         airport = if_else(origin == "EWR", "EWR", "other NY airport"))

head(all_joined_imputed)

```


### Analysis 1: Flight delay by time of day

```{r}
all_joined_imputed%>% 
  filter(origin == "EWR") %>% 
  group_by(hour, delay_dep) %>% 
  summarise(mean_delay = mean(dep_delay),
            count = n())%>% 
  ggplot()+
  geom_col(aes(x = hour, y = count, fill = factor(delay_dep, levels = c("> 1hr delay", "< 1hr delay","< 15 minutes delay",  "Early or on-time"))), position = "fill")+
  theme_classic()+
  labs(
    x = "Hour",
    y = "Proportion of delayed flights",
    title = "Flight Delay by Hour from Newark Airport",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 1))
  #scale_x_continuous(breaks=seq(1,12,1))
```


```{r}
flights_clean %>% 
  group_by(hour,origin) %>% 
  summarise(mean_delay = mean(dep_delay)) %>% 
  ggplot(aes(x = hour, y = mean_delay, colour = origin))+
  geom_point()+
  geom_line()+
  theme_classic()+
  labs(
    x = "Hour",
    y = "Average flight delay (minutes)",
    title = "Average Flight Delay by Hour from New York Airports",
    fill = "Airport"
  )+
  scale_colour_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 1))+
  scale_x_continuous(breaks = seq(0, 24, 2))
```


## Analysis 3: Weather correlation to flight delay


## Weather correlations with imputed data

```{r message=FALSE}
all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  ungroup() %>% 
  select(delayed, temp, pressure, visib, dewp, wind_speed, wind_gust, wind_dir, humid) %>% 
  ggpairs()
```


```{r message=FALSE}
all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  ungroup() %>% 
  select(dep_delay, temp, pressure, visib, dewp, wind_speed, wind_gust, wind_dir, humid) %>% 
  ggpairs()
```
```{r message=FALSE}
all_joined_imputed %>% 
  ungroup() %>% 
  select(delayed, delay_dep, wind_class, wind_direction) %>% 
  ggpairs()
```

#### Temperature
```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = temp, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Temperature (F)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```

#### Wind Gusts

```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = wind_gust, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Wind gusts (mph)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```
#### Precipitation

```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = precip, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Precipitation (inches)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```

#### Wind Speed

```{r}
all_joined_imputed %>%
  filter(origin == "EWR") %>% 
  ggplot(aes(x = wind_speed, y = delayed))+
  geom_boxplot()+
  theme_classic()+
  labs(
    x = "Wind speed (mph)",
    y = "Delayed flights",
    fill = "Departure delay"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=20))+
  theme(text = element_text(size = 20))+
  theme(plot.title = element_text(size = 20, hjust = 0.4))
```

#### Comparison of average flight delays for each weather time, and by airport


```{r}
all_joined_imputed %>% 
  group_by(wind_direction, delayed, airport) %>% 
  summarise(mean = mean(wind_speed)) %>% 
  ggplot(aes(x = wind_direction, y = mean, fill = delayed))+
  geom_col(position = "dodge")+
  theme_classic()+
  facet_wrap(~airport)+
  coord_flip()+
  labs(
    x = "Wind speed (mph)",
    y = "Delayed flights",
    fill = "Departure\ndelay",
    title = "Effect of Wind Direction on Flight Delays from New York Airports"
  )+
  scale_fill_manual(values = c("#fdae61", "#2c7bb6"))+
  theme(axis.text=element_text(size=16))+
  theme(text = element_text(size = 14))+
  theme(plot.title = element_text(size = 16, hjust = 0.4))
```


```{r}
avg_table <- all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  group_by(delayed) %>% 
  summarise(Temperature = mean(temp),
            #pressure = mean(pressure, na.rm = TRUE),
            Visibility = mean(visib, na.rm = TRUE),
            Humidity = mean(humid, na.rm = TRUE),
            #precip = mean(precip, na.rm = TRUE),
            "Wind speed" = mean(wind_speed, na.rm = TRUE),
            "Wind gust" = mean(wind_gust, na.rm = TRUE))

avg_table_pivot <- avg_table %>% 
  pivot_longer(cols = c(Temperature:"Wind gust"), names_to = "weather", values_to = "mean")

avg_table_pivot %>% 
  ggplot(aes(x = weather, y = mean, fill = delayed))+
  geom_col(position = "dodge")+
  theme_classic()+
  labs(
    y = "Mean weather",
    title = "Impact of weather on flight delays from Newark airport",
    fill = "Delayed"
  )+
  theme(axis.text=element_text(size=12))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  scale_fill_manual(values = c("#2c7bb6","#fdae61", "#abd9e9",  "#d7191c" ))
```
```{r}
avg_table_pivot %>% 
  ggplot(aes(x = mean, y = delayed))+
  geom_boxplot()+
  facet_grid(~weather)
  
```



```{r}
all_joined_imputed %>% 
  filter(delayed == TRUE) %>% 
  group_by(wind_speed, airport) %>% 
  summarise(mean = mean(dep_delay)) %>% 
  ggplot(aes(y = mean, x = wind_speed, colour = airport))+
  geom_point()+
  facet_grid(~airport)+
   theme_classic()+
  labs(
    x = "Wind speed (mph)",
    y = "Mean delay (minutes)",
    title = "Average Flight Delay with Wind Speed at New York Airports"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  theme(legend.position="none")
  
```


### Analysis: other factors
#### Airport location


```{r}
airport_table <- all_joined %>% 
  filter(origin == "EWR") %>% 
  group_by(name.y) %>% 
  summarise(count = n(),
            delayed_flights = sum(delayed),
            prop_delayed = delayed_flights/count,
            airtime = mean(air_time),
            distance = mean(distance)) %>% 
  arrange(desc(prop_delayed))
```


```{r}
# add in additional information for latitude and long missing. Easier to do this in excel. 
airport_table_latlong <- airport_table %>% 
  left_join(airports, by = c("name.y" ="name"))

write_csv(airport_table_latlong, "clean_data/airport_table_latlong.csv", append = FALSE)
```


```{r}
airport_latlong_updated <- read_csv("clean_data/airport_table_latlong_updated.csv")
```



```{r}
airport_latlong_updated %>% 
  ggplot(aes(x = count, y = prop_delayed))+
  geom_point(colour = "steelblue")+
  labs(
    x = "Number of flights Airport",
    y = "Proportion of delayed flights",
    title = "Proportion of Delayed flights compared to Total Flights by Airport"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.7))
```



```{r}
airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(delay_class = case_when(prop_delayed <= 0.31 ~ "1",
                                 prop_delayed <= 0.3738 ~ "2",
                                 prop_delayed <= 0.4143 ~ "3",
                                 prop_delayed > 0.4143 ~ "4")) 

pal <- colorFactor(
  palette = c('green', 'blue', 'orange', 'red'),
  domain = airport_latlong_updated$delay_class
)
size <- c()
```



```{r}
airport_latlong_trim <- airport_latlong_updated %>% 
  mutate(across(c(count, delayed_flights, lat, lon), as.numeric)) %>% 
  filter(!is.na(lat)) %>% 
  filter(!is.na(lon))

leaflet(airport_latlong_trim) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~lon,
                   lat = ~lat,
                   weight = 1,
                   radius = ~airport_latlong_updated$count_sqrt,
                   color = ~pal(delay_class)) %>% 
  addLegend('bottomright', pal = pal, values = airport_latlong_updated$delay_class,
            title = 'Proportion of<br>flights delayed',
            opacity = 1)
```

```{r}
airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(count_group = case_when(count < 500 ~"<500",
                                 count < 1000 ~ "<1000",
                                 count < 2000 ~ "<2000",
                                 count < 3000 ~ "<3000",
                                 count < 4000 ~ "<4000",
                                 count < 5000 ~ "<5000",
                                 count < 6000 ~ "<6000",
                                 count < 7000 ~ "<7000",
                                 count > 7000 ~ ">7000"))

airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(count_sqrt = sqrt(count)/3)
```

```{r}
airport_latlong_updated %>% 
  ggplot(aes(x = count, y = delayed_flights))+
  geom_point()+
  geom_smooth()+
  theme_classic()+
  labs(
    x = "Number of flights to Airport",
    y = "Number of delayed flights",
    title = "Number of Delayed flights compared to Total Flights by Airport"
  )+
  scale_fill_manual(values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"))+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.4))
```



```{r}
all_joined_imputed %>% 
  filter(delayed == TRUE,
         origin == "EWR") %>% 
  group_by(distance, delayed) %>% 
  summarise(mean_delay = mean(dep_delay)) %>% 
  ggplot(aes(x = distance, y = mean_delay))+
  geom_point(colour = "steelblue")+
  theme_classic()+
  labs(
    x = "Distance of flight",
    y = "Mean delay (minutes)",
    title = "Average time delay versus distance of delayed flight from Newark Airport"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  ylim(0, 100)

```

### Analysis: other factors
#### Airlines and aircraft data

```{r}
flights_weather_planes_airlines <- read_csv("clean_data/flights_weather_planes_airlines.csv")
```

```{r}
flights_weather_planes_airlines <- flights_weather_planes_airlines %>% 
  mutate(delayed = as.integer(delayed),
         delayed_1hr_more = as.integer(delayed_1hr_more))
```

```{r}
flights_weather_planes_airlines_regex <- flights_weather_planes_airlines %>% 
  mutate(name = str_remove(name, " Inc."),
         name = str_remove(name, " Co"))
```


```{r}
flights_weather_planes_airlines_regex %>% 
  group_by(name) %>% 
  summarise(count = n(),
            mean_delay = mean(dep_delay),
            perc_flights_delayed = sum(delayed)/count*100,
            perc_flights_delayed_hr = sum(delayed_1hr_more)/count*100) %>% 
  arrange(desc(perc_flights_delayed)) %>% 
  ggplot(aes(x = reorder(name, perc_flights_delayed), y = perc_flights_delayed, ))+
  geom_col(fill = "steelblue")+
  coord_flip()+
  theme_classic()+
  labs(
    y = "Percentage flights delayed",
    title = "Percentage Flight Delays by Airline"
  )+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(axis.title.y=element_blank())+
  theme(plot.title = element_text(size = 20, hjust = -0.4))
```



#### Correlation between departure delay and arrival delay

```{r}
flights_clean %>% 
  filter(delayed == "TRUE") %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  ggplot(aes(x = dep_delay, y = arr_delay))+
  geom_point()+
  geom_smooth()
```

```{r}
flight_clean %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  summarise(r_all_flights = cor(dep_delay, arr_delay))
```




```{r}
flights_weather_planes_airlines %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  filter(delayed_1hr_more == 1) %>% 
   ggplot(aes(x = dep_delay, y = arr_delay))+
  geom_point()+
  geom_smooth()
```



## Predictions for flight delays: Decision tree



```{r}
ewr_fit <- all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  ungroup() 
  

ewr_fit <- ewr_fit %>% 
  mutate(time_of_day = case_when(hour <= 6 ~ "night",
                                 hour <= 12 ~ "morning",
                                 hour <= 18 ~ "afternoon",
                                 hour <= 24 ~ "evening"),
         visibility = if_else(visib >= 5, "good", "poor"))

ewr_fit_dt <- ewr_fit %>% 
  select(delayed, month, time_of_day, distance, wind_direction, wind_class, temp, visibility, wind_gust, precip) %>% 
  mutate(delayed = as.factor(delayed)) %>% 
  na.omit()
ewr_fit_dt
```





```{r}
# test train split set up
n_data <- nrow(ewr_fit_dt)

test_index <- sample(1:n_data, size = n_data*0.2)

ewr_test <- slice(ewr_fit_dt, test_index)

ewr_train <- slice(ewr_fit_dt, -test_index)
```

```{r}
ewr_test %>% 
  janitor::tabyl(delayed)
```
```{r}
ewr_train %>% 
  janitor::tabyl(delayed)
```

```{r}
ewr_fit_tree <- rpart(
  formula = delayed ~.,
  data = ewr_train,
  method = "class"
)

rpart.plot(ewr_fit_tree,
           yesno = 2,
           fallen.leaves = TRUE,
           faclen = 6,
           digits = 2, 
           type = 4,
           extra = 101)
```
```{r}
rpart.rules(ewr_fit_tree, cover = TRUE)
```



```{r}
# add predictions to test the model
ewr_test_pred <- ewr_test %>% 
  add_predictions(ewr_fit_tree, type = "class")
```


```{r}
ewr_test_pred %>% 
  
  select(time_of_day, month, delayed, pred)  
```

```{r}
conf_mat <- ewr_test_pred %>% 
  conf_mat(truth = delayed, estimate = pred)

conf_mat
```

```{r}
accuracy <- ewr_test_pred %>% 
  accuracy(truth = delayed, estimate = pred)

accuracy
```


```{r}
confusionMatrix(ewr_test_pred$pred, ewr_test_pred$delayed)
```
The conclusion from this decision tree is that it only takes in one variable (hour) as this is the only variable that means the defined threshold for convergence. This reinforces previous analysis that there is no difference in the delays in flights with different weather. 

## Random Forest


```{r}
ewr_fit %>% 
  summary
```

```{r}
ewr_trim <- ewr_fit %>% 
  select(month, day, delayed, air_time, distance, hour, temp, wind_direction, wind_speed, wind_class, wind_gust, visibility, year.y, type, manufacturer, model, engines, seats, engine, name.x, lat, lon, alt, tz) %>% 
  na.omit()


ewr_trim 
```


```{r}
rf_classifier <- ranger(delayed ~ .,
                       data = ewr_trim, 
                       importance = "impurity",
                       num.trees = 1000,
                       mtry = 2, 
                       min.node.size = 5)

rf_classifier
```

```{r}
sort(importance(rf_classifier))
```

From this we see that hour, airtime, day, minute and year.y(ie year of plane) are the most important variables, followed by wind_gust, wind_speed, month, wind_direction, temperature, model, lat, distance, altitude, longitude, seats, name.x, manufacturer, visibility, tz, engine, wind_class, type and engines. 




## Prediction of flight delay using logistical regression

BAsed on the delayed variable (True/false) and single variable for hour. 

```{r}
# set up the dataset

all_joined_model <- all_joined_imputed %>% 
  select(date, time, delayed, hour)
```


```{r}
n_data <- nrow(all_joined_model)

delay_index <- sample(1:n_data, size = n_data*0.2)

delay_test <- slice(all_joined_model, test_index)

delay_train <- slice(all_joined_model, -test_index)
```

```{r}
delay_test %>% 
  janitor::tabyl(delayed)
```
```{r}
delay_train%>% 
  janitor::tabyl(delayed)
```


```{r}
delayed_1log_reg <- glm(delayed ~ hour, 
                             data = delay_train,
                             family = binomial(link = "logit"))

delayed_1log_reg
```


```{r}
test_1log_reg <- glm(delayed ~ hour,
                     data = delay_test, 
                     family = binomial(link = "logit"))
```

Interpretation

ln(odds_delayed(x)) = b0 + b1 * x

in this case:
ln(odds_delayed(x)) = -2.1996 + (0.1242 * x)

$$              
\ln({odds_{delayed}(x))}=-2.1996 + (0.1242 * (x))

$$

```{r}
# add predictions
predict_log <- tibble(hour = seq(0, 24, 0.5)) %>% 
  add_predictions(delayed_1log_reg, 
                  type = "response")

predict_log

delay_train %>% 
  ggplot(aes(x = hour, y = as.integer(delayed))) +
  geom_jitter(position = position_jitter(h = 0.1), shape = 1)+
  geom_line(data = predict_log,
            aes(y = pred),
            col = "red")+
  labs(title = "Probability of Flight Delay Based on Hour",
       y = "Probability",
       x = "Hour")+
  theme_classic()+
  theme(axis.text=element_text(size=14))+
  theme(text = element_text(size = 16))+
  theme(plot.title = element_text(size = 16, hjust = 0.2))+
  scale_x_continuous(breaks=seq(0,24,2))
```

```{r}
tidy_out <- clean_names(tidy(delayed_1log_reg))
tidy_out
```

```{r}
tidy_test <- clean_names(tidy(test_1log_reg))
tidy_test
```


```{r}
summary(delayed_1log_reg)
```


```{r}
delay_train_with_1pred <- delay_train %>% 
  add_predictions(delayed_1log_reg, type = "response")

delay_train_1pred %>% 
  ggplot(aes(x = hour, y = pred, colour = delayed))+
  geom_line()
```

```{r}
test_with_1pred <- delay_test %>% 
  add_predictions(delayed_1log_reg, type = "response")

test_with_1pred %>% 
   ggplot(aes(x = hour, y = pred, colour = delayed))+
  geom_line()
```


Neither threshold is particularly good!

```{r}
roc_obj_1pred <- delay_train_with_1pred %>% 
  roc(response = delayed,
      predictor = pred)

auc(roc_obj_1pred)
```
```{r}
roc_obj_test1pred <- test_with_1pred %>% 
  roc(response = delayed,
      predictor = pred)

auc(roc_obj_test1pred)
```






```{r}
roc_curve <- ggroc(data = list(
  mod1 = roc_obj_1pred,
  mod2 = roc_obj_test1pred),
                   legacy.axes = TRUE)+
  coord_fixed()+
  labs(x = "FPR = 1 - TNR = 1 - specificity",
          y = "TPR = sensitivity")

roc_curve
```
The simple logistical model with a single predictor (hour of day) shows a better than guess AUC of 0.66. It could be improved by the addition of more predictors, however the p-values show that the variable is significant and can be used to determine the probability (and the odds) of a flight being delayed, based on a single predictor of hour of flight. 

## Conclusions

There was very little evidence of weather impacting at all on flight delays. Almost all of the weather variables showed little difference for delayed and on-time flights. This was confirmed using correlation plots and also with decision trees, which only picked out 'Hour' as a significant variable. 

The largest effect on flight delays was seen with time of day. There was a relation between increasing hour of the day and increasing chance of flight being delayed, and for the delay also to increase in time. THere was an additional slight variation in flight delays with month, which could be due to a higher number of flights during the busy summer time, compared to Autumn. There is a potential to stagger flights during the day to mitigate the knock on effects of delays on later flights. 

There was a big difference in percentage of flight delays with different airlines, South Western showed 45% flights delayed, American airlines 30% and Hawaiwan Airlines 20%. This could be investigated with the airline companies, as delays are costly. 

## Future work

It would be very interesting to investigate with the airlines the different types of planes, ages, engines, types which might be correlated to the delay times and might be a useful investment for companies. In addition, it would be useful to compare to the other airports further too, the limits of this project didn't allow. 

Furthermore, the model could be improved by the addition of more variables, especially those with ranked importance from the random forest. This may result in an improved model, with better capability to predict future delays in flights from Newark Airport. 


