---
title: "R Notebook"
output: html_notebook
---


# Flight delays project

```{r}
library(tidyverse)
library(here)
library(GGally)
library(modelr)
library(leaflet)
library(rpart)
library(rpart.plot)
library(yardstick)
library(caret)
library(ranger)
```

```{r}
flights_clean <- read_csv(here("clean_data/flight_clean.csv"))
```

## Flight delays

1. How many flights are delayed?
2. By how much are flights delayed?



```{r}
flights_clean %>% 
  filter(origin == "EWR") %>% 
  group_by(delay_dep) %>% 
  count() %>% 
  ggplot()+
  geom_col(aes(x = reorder(delay_dep, n), y = n), fill = "steelblue")+
  coord_flip()+
  theme_classic()+
  labs(
    x = "Flight delay",
    y = "Number of flights",
    title = "Number of flights delayed from Newark Airport in 2017"
  )
```

```{r}
flights_clean %>% 
  group_by(delay_dep, origin) %>% 
  summarise(count = n()) %>% 
  ggplot()+
  geom_col(aes(x = origin, y = count, fill = delay_dep), position = "fill")+
  theme_classic()+
  labs(
    x = "Flight delay",
    y = "Number of flights",
    title = "Comparison of Flight Delays from New York Airports in 2017"
  )
```

```{r}
flights_clean %>% 
  filter(delay_dep %in% c( "< 1hr delay", "> 1hr delay"),
         origin == "EWR") %>% 
  ggplot(aes(x = dep_delay))+
  geom_histogram(bin = 50)+
  xlim(0, 1000)
```

? Could do a skewness calculation?


```{r}
flights_clean %>% 
  filter(delay_arr %in% c( "< 1hr delay", "> 1hr delay"),
         origin == "EWR") %>% 
  ggplot(aes(x = arr_delay))+
  geom_histogram(bin = 50)+
  xlim(0, 1000)
```


```{r}
flights_clean %>% 
  group_by(origin, delay_dep) %>% 
  summarise(count = n(),
            proportion = n()/nrow(flights_clean))
```


```{r}
flights_clean %>% 
  group_by(delay_dep, origin) %>% 
  summarise(count = n(),
            percentage = n()/nrow(flights_clean)*100) %>% 
  ggplot()+
  geom_col(aes(x = delay_dep, y = percentage))+
  theme_classic()+
  labs(
    x = "Flight delay",
    y = "Number of flights",
    title = "Number of flights delayed from Newark Airport in 2017"
  )+
  facet_grid(~origin)
```



Delay proportion by month to check for seasonal variation

```{r}
flights_clean %>% 
  filter(origin == "EWR") %>% 
  group_by(month, delay_dep) %>% 
  count() %>% 
  ggplot(aes(x = month, y = n, fill = delay_dep))+
  geom_col()
```



```{r}
flight_clean %>% 
  filter(origin == "EWR",
         delay_dep != "Early or on-time") %>% 
  drop_na(arr_delay) %>% 
  group_by(month) %>% 
  summarise(mean_delay = mean(dep_delay),
            mean_arr_delay = mean(arr_delay),
            count = n()) %>% 
  ggplot(aes(x = month, y = mean_delay))+
  geom_col()
```


```{r}
flights_clean %>% 
   filter(origin == "EWR") %>% 
  drop_na(arr_delay) %>% 
  group_by(month) %>% 
  summarise(mean_delay = mean(dep_delay),
            mean_arr_delay = mean(arr_delay),
            count = n()) %>% 
  ggplot(aes(x = month, y = mean_arr_delay, colour = count))+
  geom_point(size = 3)
```


```{r}
flight_clean %>% 
  filter(origin == "EWR",
         delay_dep != "Early or on-time") %>% 
  group_by(month) %>% 
  summarise(mean_delay = mean(dep_delay),
            mean_arr_delay = mean(arr_delay, na.rm = TRUE),
            count = n()) %>% 
  ggplot(aes(x = month, y = mean_arr_delay))+
  geom_col()
```
```{r}
flights_clean %>% 
  group_by(hour, delayed) %>% 
  count() %>% 
  ggplot(aes(x = hour, y = n, fill = delayed, position = "dodge"))+
  geom_col(position = "dodge")
```
```{r}
flights_clean %>% 
  group_by(hour) %>% 
  summarise(mean_delay = mean(dep_delay)) %>% 
  ggplot(aes(x = hour, y = mean_delay))+
  geom_point()
```

```{r}
flights_clean %>% 
  ggplot(aes(x = sched_dep_time, y = dep_delay))+
  geom_point()
```

```{r}
flights_clean %>% 
  ggplot(aes(x = sched_arr_time, y = arr_delay))+
  geom_point()
```


```{r}
flight_clean %>% 
  filter(origin == "EWR") %>% 
  drop_na(delay_diff) %>% 
  group_by(month) %>% 
  summarise(month_diff = mean(delay_diff)) %>% 
  ggplot(aes(x = month, y = month_diff))+
  geom_col()
```


Filter out the days with very high delayed flights

```{r}
flights_clean %>% 
  filter(delay_dep == "> 1hr delay") %>% 
  group_by(date) %>% 
  count() %>% 
  arrange(desc(n))
```


```{r}
flights_clean %>% 
  group_by(origin) %>% 
  count()
```

Airport comparisons 

```{r}
flight_clean %>% 
  group_by(origin, delay_dep) %>% 
  summarise(count = n()) %>%
  ggplot()+
  geom_col(aes(x = delay_dep, y = count))+
  theme_classic()+
  labs(
    x = "Flight delay",
    y = "Number of flights",
    title = "Number of flights delayed from Newark Airport in 2017"
  )+
  facet_grid(~origin)
```
WEather correlations

```{r}
newark_flights_weather <- read_csv("clean_data/newark_flights_weather.csv")
flights_weather <- read_csv("clean_data/flights_weather.csv")
```

```{r}
newark_flights_weather %>% 
  glimpse
```




```{r}
newark_flights_weather %>% 
  ggplot(aes(x = temp, y = delayed))+
  geom_boxplot()
```

```{r}
newark_flights_weather %>% 
  ggplot(aes(x = hour, y = delayed))+
  geom_boxplot()
```



```{r message=FALSE}
newark_flights_weather %>% 
  select(delayed, wind_speed, wind_gust, temp, visib) %>% 
  ggpairs()
```

```{r}
weather_clean %>% 
  filter(time_hour == "2017-01-01 09:00:00")
```


```{r}
flights_weather %>% 
  glimpse()
```
```{r}
flights_weather <- flights_weather %>% 
  mutate(wind_class = case_when(wind_speed <= 24 ~"Breeze",
                                wind_speed <= 31 ~"Strong breeze",
                                wind_speed <= 38 ~ "Near gale",
                                wind_speed > 38 ~ "Gale"),
         .after = wind_speed)
```



```{r}
group1<- flights_weather %>% 
  select(month, day, hour, minute, dep_delay, delay_dep, delayed, delayed_1hr_more)
```


```{r}
group2 <- flights_weather %>% 
  select(delayed, delayed_1hr_more, arr_time, arr_delay, delay_diff, carrier, origin, time_hour, distance)
```

```{r}
group3 <- flights_weather %>% 
  select(delayed, delayed_1hr_more, delay_dep, air_time, distance, wind_class, temp, wind_dir, wind_speed, visib )
```

```{r message=FALSE}
group1 %>% 
  ggpairs()
```

```{r message=FALSE}
group2 %>% 
  ggpairs()
```

```{r message=FALSE}
group3 %>% 
  ggpairs()
```

```{r}
flights_weather %>% 
  group_by(delayed_1hr_more) %>% 
  count()
```



wind speed 

```{r}
weather_clean %>% 
  summary
```

```{r}
flights_weather <- flights_weather %>% 
  mutate(wind_class = case_when(wind_speed <= 24 ~"Breeze",
                                wind_speed <= 31 ~"Strong breeze",
                                wind_speed <= 38 ~ "Near gale",
                                wind_speed > 38 ~ "Gale"),
         .after = wind_speed)
```

```{r}
flights_weather %>% 
  filter(is.na(wind_class))
```

# plane data for delays too

```{r}
flights_weather_planes <- read_csv("clean_data/flights_weather_planes.csv")
```

```{r}
flights_weather_planes %>% 
  
```

```{r}
group4 <- flights_weather_planes %>% 
  select(delayed, delayed_1hr_more, year.y, type, engine, engines, seats, speed)
```

```{r}
flights_weather_planes %>% 
  group_by(manufacturer) %>% 
  count() %>% 
  arrange(desc(n))
```

```{r}
flights_weather_planes %>% 
  group_by(model) %>% 
  count() %>% 
  arrange(desc(n))
```


```{r}
group4 %>% 
  ggpairs()
```

# airlines data

```{r}
flights_weather_planes_airlines <- read_csv("clean_data/flights_weather_planes_airlines.csv")
```


```{r}
flights_weather_planes_airlines %>% 
  glimpse()
```

```{r}
group5 <- flights_weather_planes_airlines %>% 
  select(delayed, delayed_1hr_more, delay_dep, name, engine, seats, type, year.y) 
```

```{r message=FALSE}
group5 %>% 
  ggpairs()
```
```{r}
flights_weather_planes_airlines <- flights_weather_planes_airlines %>% 
  mutate(delayed = as.integer(delayed),
         delayed_1hr_more = as.integer(delayed_1hr_more))
```



```{r}
flights_weather_planes_airlines %>% 
  group_by(name) %>% 
  summarise(count = n(),
            mean_delay = mean(dep_delay),
            perc_flights_delayed = sum(delayed)/count*100,
            perc_flights_delayed_hr = sum(delayed_1hr_more)/count*100) %>% 
  arrange(desc(perc_flights_delayed)) %>% 
  ggplot(aes(x = reorder(name, mean_delay), y = mean_delay, ))+
  geom_col()+
  coord_flip()
```

```{r}
flights_weather_planes_airlines %>% 
  group_by(model) %>% 
  summarise(count = n(),
            mean_delay = mean(dep_delay),
            delay_per_plane = count/mean_delay,
            perc_flights_delayed = sum(delayed)/count*100,
            perc_flights_delayed_hr = sum(delayed_1hr_more)/count*100) %>% 
  arrange(desc(perc_flights_delayed_hr)) 
```

```{r}
flights_weather_planes_airlines %>% 
  group_by(manufacturer) %>% 
  summarise(count = n(),
            mean_delay = mean(dep_delay),
            perc_flights_delayed = sum(delayed)/count*100,
            perc_flights_delayed_hr = sum(delayed_1hr_more)/count*100) %>% 
  arrange(desc(perc_flights_delayed)) 
```

```{r}
delay_table <- flights_weather_planes_airlines %>% 
  group_by(time_hour) %>% 
  summarise(count = n(),
            flights_delayed = sum(delayed),
            flights_delayed_more1hr = sum(delayed_1hr_more),
            perc_delayed = flights_delayed/count*100,
            perc_delay_more_1hr = sum(delayed_1hr_more)/count*100) 
            
```

join delay_table to weather

```{r}
delay_table_weather <- delay_table %>% 
  left_join(weather_clean, by = "time_hour")
```

```{r message=FALSE}
delay_table_weather %>% 
  select(time_hour, count, perc_delayed, perc_delay_more_1hr, origin, temp, wind_speed, visib, wind_dir, wind_gust) %>% 
  ggpairs()
```

Correlation between departure delay and arrival delay

```{r}
flights_clean %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  ggplot(aes(x = dep_delay, y = arr_delay))+
  geom_point()+
  geom_smooth()
```

```{r}
flights_clean %>% 
  filter(delayed == "TRUE") %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  ggplot(aes(x = dep_delay, y = arr_delay))+
  geom_point()+
  geom_smooth()
```



```{r}
flight_clean %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  summarise(r_all_flights = cor(dep_delay, arr_delay))
```



```{r}
flight_clean %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  filter(delayed == TRUE) %>% 
  summarise(r_delayed_flights = cor(dep_delay, arr_delay))
```

```{r}
flight_clean %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  filter(delayed == TRUE) %>% 
   ggplot(aes(x = dep_delay, y = arr_delay))+
  geom_point()+
  geom_smooth()+
  theme_classic()+
  labs(
    title = "Correlation between departure and arrival times of delayed flights from EWR",
    x = "Departure delay (minutes)",
    y = "Arrival delay (minutes)",
    subtitle = "R = 0.961"
  )
  
```


```{r}
flights_weather_planes_airlines %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  filter(delayed_1hr_more == 1) %>% 
   ggplot(aes(x = dep_delay, y = arr_delay))+
  geom_point()+
  geom_smooth()
```

```{r}
flight_clean %>% 
  drop_na(arr_delay) %>% 
  drop_na(dep_delay) %>% 
  filter(delayed_1hr_more == TRUE) %>% 
  summarise(cor(dep_delay, arr_delay))
```
Model building

simple linear regression for dep_delay verses wind_speed

```{r}
flights_weather %>% 
  group_by(wind_class) %>% 
  summarise(mean_delay = mean(dep_delay),
            mean_wind_speed = mean(wind_speed),
            count = n()) %>% 
  ggplot(aes(x = mean_wind_speed, y = mean_delay, size = count, colour = wind_class))+
  geom_point()
```


```{r}
flights_weather %>% 
  filter(wind_class == "Strong breeze",
         month == 5)
```


```{r}
flights_weather %>% 
  ggplot(aes(y = dep_delay, x = wind_speed, colour = wind_class))+
  geom_jitter()
```

```{r}
model_slr1 <- lm(formula = dep_delay ~ wind_gust + wind_speed + temp + visib, data = flights_weather)

model_slr1
```
```{r}
flights_weather_planes_airlines %>% 
  distinct(visib)
```


```{r}
flights_weather_planes_airlines_trim <- flights_weather_planes_airlines %>% 
  select(-c(year.x, date, dep_time, sched_dep_time, arr_time, sched_arr_time, flight, tailnum, time_hour, delayed, delayed_1hr_more, engines)) %>% 
  glimpse()

model_weather <- (lm(dep_delay ~ wind_dir + wind_speed + wind_gust + precip + pressure + visib, data = flights_weather_planes_airlines_trim))

model_weather
```

```{r}
flights_weather_planes_airlines_trim %>% 
  distinct(name)
```


All datasets joined

```{r}
all_joined <- read_csv("clean_data/all_joined.csv")
```

```{r}
airport_table <- all_joined %>% 
  filter(origin == "EWR") %>% 
  group_by(name.y) %>% 
  summarise(count = n(),
            delayed_flights = sum(delayed),
            prop_delayed = delayed_flights/count) %>% 
  arrange(desc(prop_delayed))
```

```{r}
airports <- read_csv("raw_data/airports.csv")
```

```{r}
airport_table_latlong <- airport_table %>% 
  left_join(airports, by = c("name.y" ="name"))

write_csv(airport_table_latlong, "clean_data/airport_table_latlong.csv", append = FALSE)
```


```{r}
airport_latlong_updated <- read_csv("clean_data/airport_table_latlong_updated.csv")
```


```{r}
all_joined %>% 
  filter(is.na(name.y)) %>% 
  distinct(dest)
```

```{r}
airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(delay_class = case_when(prop_delayed <= 0.31 ~ "1",
                                 prop_delayed <= 0.3738 ~ "2",
                                 prop_delayed <= 0.4143 ~ "3",
                                 prop_delayed > 0.4143 ~ "4")) 

pal <- colorFactor(
  palette = c('green', 'blue', 'orange', 'red'),
  domain = airport_latlong_updated$delay_class
)
size <- c()
```

```{r}
airport_latlong_updated %>% 
  mutate(count_group = case_when(count < 500 ~"<500",
                                 count < 1000 ~ "<1000",
                                 count < 2000 ~ "<2000",
                                 count < 3000 ~ "<3000",
                                 count < 4000 ~ "<4000",
                                 count < 5000 ~ "<5000",
                                 count < 6000 ~ "<6000",
                                 count < 7000 ~ "<7000",
                                 count > 7000 ~ ">7000"))

airport_latlong_updated <- airport_latlong_updated %>% 
  mutate(count_sqrt = sqrt(count)/3)
```



```{r}
airport_latlong_trim <- airport_latlong_updated %>% 
  mutate(across(c(count, delayed_flights, lat, lon), as.numeric)) %>% 
  filter(!is.na(lat)) %>% 
  filter(!is.na(lon))

leaflet(airport_latlong_trim) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~lon,
                   lat = ~lat,
                   weight = 1,
                   radius = ~airport_latlong_updated$count_sqrt,
                   color = ~pal(delay_class)) %>% 
  addLegend('bottomright', pal = pal, values = airport_latlong_updated$delay_class,
            title = 'Quartile groups<br>proportion of<br>flights delayed',
            opacity = 1)
```



## Decision trees

First need to impute data into the NAs for wind speed and wind gusts. Want to swap NAs for the mean speed for that day. 

```{r}
all_joined %>% 
  summary
```


```{r}
all_joined_imputed <- all_joined %>% 
  group_by(date, origin) %>% 
  mutate(across(.cols = temp:wind_speed, .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)),
         across(.cols = wind_gust:visib, .fns = ~if_else(is.na(.x), mean(.x, na.rm = TRUE), .x)))
```

```{r}
all_joined_imputed <- all_joined_imputed %>% 
  group_by(month, origin) %>% 
  mutate(temp = if_else(is.na(temp), mean(temp, na.rm = TRUE), temp))
```


```{r}
all_joined_imputed <- all_joined_imputed %>% 
  mutate(wind_class = case_when(wind_speed <= 24 ~"Breeze",
                                wind_speed <= 31 ~"Strong breeze",
                                wind_speed <= 38 ~ "Near gale",
                                wind_speed > 38 ~ "Gale"),
         .after = wind_speed)
```


```{r}
all_joined_imputed <- all_joined_imputed %>% 
  mutate(wind_direction = case_when(wind_dir <= 23 ~ "North",
                                    wind_dir <= 68 ~ "Northeast",
                                    wind_dir <= 113 ~ "East",
                                    wind_dir <= 158 ~ "Southeast",
                                    wind_dir <= 203 ~ "south",
                                    wind_dir <= 248 ~ "southwest",
                                    wind_dir <= 293 ~ "West",
                                    wind_dir <= 336 ~ "northwest",
                                    wind_dir > 336 ~ "north"),
         .after = wind_dir,
         delayed = as.logical(delayed))
head(all_joined_imputed)
```

### Correlation plots with new imputed weather data for Newark airport
```{r}
ewr_all_joined <- all_joined_imputed %>%
   filter(origin == "EWR")

ewr_weather %>% 
  select(month, dep_delay, wind_direction, wind_speed, wind_class, temp, visib)
 
```

```{r message=FALSE}
ewr_weather %>% 
  ggpairs()
```

```{r}
all_joined_trim <- all_joined_imputed %>% 
  select(date, origin, month, day, delayed, delay_dep, delay_arr, air_time, distance, hour, temp, dewp, humid, wind_direction, wind_class, visib, type)
```


```{r message=FALSE}
all_joined_trim %>% 
  select(date, origin, month, hour, day, delayed, delay_arr, air_time, distance) %>% 
  ggpairs()
```

```{r message=FALSE}
all_joined_trim %>% 
  select(delayed, hour, temp, wind_direction, wind_class, visib) %>% 
  ggpairs()
```
```{r}
ewr_all_joined_trim
```


```{r}
all_joined_imputed %>%
  ggplot()+
  aes(x = month, y = delayed)+
  geom_boxplot()
```


## Decision Tree!

```{r}
ewr_fit <- all_joined_imputed %>% 
  filter(origin == "EWR") %>% 
  ungroup() 
  

ewr_fit <- ewr_fit %>% 
  mutate(time_of_day = case_when(hour <= 6 ~ "night",
                                 hour <= 12 ~ "morning",
                                 hour <= 18 ~ "afternoon",
                                 hour <= 24 ~ "evening"),
         visibility = if_else(visib >= 5, "good", "poor"))

ewr_fit_dt <- ewr_fit %>% 
  select(delayed, month, time_of_day, distance, wind_direction, wind_class, temp, visibility, wind_gust) %>% 
  mutate(delayed = as.factor(delayed)) %>% 
  na.omit()
ewr_fit_dt
```

test train split set up

```{r}
n_data <- nrow(ewr_fit_dt)

test_index <- sample(1:n_data, size = n_data*0.2)

ewr_test <- slice(ewr_fit_dt, test_index)

ewr_train <- slice(ewr_fit_dt, -test_index)
```

```{r}
ewr_test %>% 
  janitor::tabyl(delayed)
```
```{r}
ewr_train %>% 
  janitor::tabyl(delayed)
```

```{r}
ewr_fit_tree <- rpart(
  formula = delayed ~.,
  data = ewr_train,
  method = "class"
)

rpart.plot(ewr_fit_tree,
           yesno = 2,
           fallen.leaves = TRUE,
           faclen = 6,
           digits = 2, 
           type = 4,
           extra = 101)
```
```{r}
rpart.rules(ewr_fit_tree, cover = TRUE)
```



```{r}
# add predictions to test the model
ewr_test_pred <- ewr_test %>% 
  add_predictions(ewr_fit_tree, type = "class")
```


```{r}
ewr_test_pred %>% 
  
  select(time_of_day, month, delayed, pred)  
```

```{r}
conf_mat <- ewr_test_pred %>% 
  conf_mat(truth = delayed, estimate = pred)

conf_mat
```

```{r}
accuracy <- ewr_test_pred %>% 
  accuracy(truth = delayed, estimate = pred)

accuracy
```
```{r}
ewr_test_pred %>% 
  sensitivity(truth = delayed, estimate = pred)
```

```{r}
ewr_test_pred %>% 
  specificity(truth = delayed, estimate = pred)
```

```{r}
confusionMatrix(ewr_test_pred$pred, ewr_test_pred$delayed)
```


## Random Forest


```{r}
ewr_fit %>% 
  summary
```

```{r}
ewr_trim <- ewr_fit %>% 
  select(month, day, delayed, air_time, distance, hour, minute, temp, wind_direction, wind_speed, wind_class, wind_gust, visibility, year.y, type, manufacturer, model, engines, seats, engine, name.x, lat, lon, alt, tz) %>% 
  na.omit()


ewr_trim %>% 
  summary
```


```{r}
rf_classifier <- ranger(delayed ~ .,
                       data = ewr_trim, 
                       importance = "impurity",
                       num.trees = 1000,
                       mtry = 2, 
                       min.node.size = 5)

rf_classifier
```

```{r}
sort(importance(rf_classifier))
```

From this we see that hour, airtime, day, minute and year.y(ie year of plane) are the most important variables, followed by wind_gust, wind_speed, month, wind_direction, temperature, model, lat, distance, altitude, longitude, seats, name.x, manufacturer, visibility, tz, engine, wind_class, type and engines. 



rf_preds <- thrones_test %>% 
  mutate(pred = predict(rf_classifer, thrones_test)$predictions)

rf_preds %>% 
  select(dth_flag, pred)
  
```{r}
rf_preds <- ewr_test %>% 
  mutate(pred = predict(rf_classifier, data = ewr_test$predictions))

rf_preds %>% 
  select(delayed, preds)
```
  



```{r}
confusionMatrix(rf_classifier$pred, rf_classifier$delayed)
```

